# Bios 301: Assignment 1 #

*Due Monday, 1 October, 12:00 PM*

```{r setup, echo=FALSE}
render_markdown(strict=TRUE)
opts_chunk$set(comment=NA)
```

50 points total.

Submit a single knitr (either `.rnw` or `.rmd`) file, along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Raw R code/output or word processor files are not acceptable. 

1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)
    
        ```{r}
        cancer.df <- read.csv("../datasets/cancer.csv")
        ```
    
    2. Determine the number of rows and columns in the data frame. (2)
    
        ```{r}
        dim(cancer.df)
        ```
    
    3. Extract the names of the columns in `cancer.df`. (2)
    
        ```{r}
        names(cancer.df)
        ```
    
    4. Report the value of the 3000th row in column 6. (2)
    
        ```{r}
        cancer.df[3000,6]
        ```
    
    5. Report the contents of the 172nd row. (2)
    
        ```{r}
        cancer.df[172,]
        ```
    
    6. Create a new column that is the incidence *rate* (per 100,000) for each row.(3)
    
        ```{r}
        cancer.df$incidence_rate <- 100000 * cancer.df$incidence / cancer.df$population
        ```
    
    7. How many subgroups (rows) have a zero incidence rate? (2)
    
        ```{r}
        sum(cancer.df$incidence==0)
        ```
    
    8. Find the subgroup with the highest incidence rate.(3)
    
        There are several ways of doing this; here is one:
    
        ```{r}
        cancer.df[which(cancer.df$incidence_rate==max(cancer.df$incidence_rate)),]
        ```


2. **Data types** (14 points)

    1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (6 points)
           
            max(x)
            sort(x)
            sum(x)
            
        Only `sum` produces an error, because the function requires numeric arguments. `sort` and `max` produce unexpected results, but if you consider that we are passing strings, it is pretty reasonable -- they are returning the alphabetical order and the last item in the order, respectively.
           
    2. For the next two commands, either explain their results, or why they should produce errors. (4 points)
    
            y <- c("5",7,12)
            y[2] + y[3]
            
        Vectors must be of the same mode, so `c()` casts everything to characters. Hence, we end up adding characters, which does not work.
        
    3. For the next two commands, either explain their results, or why they should produce errors. (4 points)
    
            z <- data.frame(z1="5",z2=7,z3=12)
            z[1,2] + z[1,3]
            
        Data frames allow columns of different modes, so 7 and 12 are not cast to strings, and so can be added.
            
3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    
        ```{r}
        c(seq(1,8),seq(7,1))
        ```
    
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
    
        ```{r}
        rep(1:5, 1:5)
        ```
    
    3. $\left({
    \begin{array}{c}
      0 & 1 & 1  \cr
      0 & 0 & 1  \cr
      1 & 1 & 0  \cr
    \end{array}
    }\right)$
    
        Here are two ways. First, the concise but tricky way:
        
        ```{r}
        (diag(3)==0) * 1
        ```
        
        and second, the straightforward way:
        
        ```{r}
        matrix(rep(1,9),nr=3) - diag(3)
        ```
    
    4. $\left({
    \begin{array}{c}
      0 & 2 & 3  \cr
      0 & 5 & 0  \cr
      7 & 0 & 0  \cr
    \end{array}
    }\right)$
    
        Here is one solution. I will accept any reasonable answer here, since this is tricky:
        
        ```{r}
        matrix(1:9 * (1:9 %in% c(2,3,5,7)), nr=3, byrow=T)
        ```
    
4. **Basic programming** Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x_i$. Write an R program to calculate $h(x,n)$ using a `for` loop. (6 points)

    ```{r}
    h <- function(x,n) {
        tot <- 1
        for (i in 1:n) tot <- tot + x^i
        return(tot)
    }
    ```


<!-- Mathjax -->
<script type="text/x-mathjax-config">
  MathJax.Hub.Config({
    tex2jax: {
      displayMath: [ ['$$','$$'], ["\\[","\\]"] ],
      inlineMath: [ ['$','$'], ["\\(","\\)"] ],
      processEscapes: true
    }
  });
</script>
<script type="text/javascript"
    src="http://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML">
</script>