---
title: 'Bios 301: Assignment 3'
output: pdf_document
---

*Due Tuesday, 11 November, 1:00 PM*

$5^{n=day}$ points taken off for each day late.

50 points total.

Submit a single knitr file (named `homework3.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

### Question 1 ###

**15 points**

**20 points**

The game of craps is played as follows. First, you roll two six-sided dice; let x be the sum of the dice on the first roll. If x = 7 or 11 you win, otherwise you keep rolling until either you get x again, in which case you also win, or until you get a 7 or 11, in which case you lose.

Write a program to simulate a game of craps. You can use the following snippet of code to simulate the roll of two (fair) dice:

```{r}
    x <- sum(ceiling(6*runif(2)))
```

1. The instructor should be able to easily import and run your program (function), and obtain output that clearly shows how the game progressed.  Set the RNG seed with `set.seed(100)` and show the output of three games. (15 points)

1. Find a seed that will win ten straight games.  Consider adding an argument to your function that disables output. Show the output of the ten games. (5 points)

### Question 2 ###

**20 points**

Code a function that does golden section search, and use this function to find all of the global maxima on the following function:

$$f(x) = \left\{
  \begin{array}{ll}0 & \text{if } x=0 \\
  |x| \log\left(\frac{|x|}{2}\right)e^{-|x|} & \text{otherwise}
  \end{array}
\right.
$$

on the interval [-10, 10].

To get an idea of what the function looks like, it might be helpful to plot it.

### Question 3 ###

**10 points**

Obtain the code for using Newton's Method to estimate logistic regression parameters (`logistic.r`) and modify it to predict `death` from `weight`, `hemoglobin` and `cd4baseline` in the HAART dataset. Use complete cases only. Report the estimates for each parameter, including the intercept.

Note: The original script `logistic_debug.r` is in the exercises folder.  It needs modification, specifically, the logistic function should be defined:

```{r}
logistic <- function(x) 1 / (1 + exp(-x))
```

### Question 4 ###

**5 bonus points**

Import the `addr.txt` file from the GitHub repository.  This file contains a listing of names and addresses (thanks google).  Parse each line to create a data.frame with the following columns: lastname, firstname, streetno, streetname, city, state, zip.  Keep middle initials or abbreviated names in the firstname column.  Print out the entire data.frame.
