---
title: "Writing and Using Functions"
output: pdf_document
---

## Functions

Functions provide the foundation for building large programs, allowing complex algorithms to be efficiently structured.

Just as data structures combine related values in a single object, functions combine related commands and variables in a single object.

Functions are just computer implementations of mathematical functions:

* they encapsulate a strict set of rules that take some **input** and use it to calculate one or more **output** values.

We have already seen a suite of built-in functions in R:

* `exp()`
* `log()`
* `ifelse()`
* `which()`
* `max()`

---

## Calling Functions

Functions are used by *calling* them, which involves passing zero or more arguments to the function by enclosing them in parentheses after the function name.

**Example:** the linear model (`lm`) function:

```{r}
data(iris)
iris_model <- lm(Sepal.Length ~ Sepal.Width, data=iris)
summary(iris_model)
```

---

## Writing Functions

We write functions because we often find ourselves going through the same sequence of steps in the R shell. It is more efficient to take a sequence of commands and bind it together into an integrated procedure, so that then we can think about the function as a whole, rather than the individual steps.

It also reduces error, because by invoking the same function every time, we don’t have to worry about missing a step, or about whether code will behave consistently each time we use it.

We can write (or *declare*) functions using a function  ... called `function`:

```{r}
my_function <- function(first_argument, second_argument, ...) {
  # do some stuff
  return(return_value)
}
```

---

## Writing Functions

An example of a very simple (and redundant) function is one which returns the absolute value of its argument[1]:

```{r}
absval <- function(value) {
    if (value<0) {
        -value
    } else {
        value
    }
}
```

By default, a function returns the last evaluation, but it is often clearer to explicitly use the `return` statement:

`return(value)`

Function's inputs (arguments) and outputs (return values) are called *interfaces*.

1. *Can also be written as:*

```{r}
absval <- function(value) if (value<0) -value else value
```

---

## Exercise

### Vectorize the `absval` function

## Presenter Notes

Hint: use vector indexing!

---

## Operators are Functions

Even the simple arithmetic operators are functions, which are simply allowed to be called differently so that they can be used more naturally.

```{r}
'+'(4,5)
```

Users can write custom operators. In order that they may be used like an operator, the name must be enclosed within percentage symbols (%). For example, suppose we wanted an operator that divided the value on the left by the square root of the value on the left:

```{r}
"%!%" = function(x,y) x/sqrt(y)
6 %!% 7
```

---

## Default Argument Values

We sometimes want to make arguments optional, or to have a default value when it is an extremely common choice. We can use *named arguments* for this:

```{r}
psi <- function(x, c=1) {
  loss <- ifelse(x^2 > c^2, c*abs(x), x^2)
  return(loss)
}
z <- 0.5
psi(z)
```

The above is a loss function for robust regression. The point at which we switch from using absolute value loss and squared loss is set to 1 if it is not specified by the user.

```{r}
identical(psi(z,c=1), psi(z))
```

Arguments can go in any order when explicitly named:

```{r}
identical(psi(x=z,c=2), psi(c=2,x=z))
```

---

## Example: Pareto Quantiles

Let's code a real (non-trivial) function in R. The **Pareto distribution** is a continuous random variable that is often used to describe the "80/20" rule.

$$f(x|a,c) = ca^cx^{-(c+1)}$$
$$x \ge a$$
$$a,c > 0$$

The cumulative distribution function (CDF) is:

$$F(x|a,c) = 1 - \left(\frac{x}{a}\right)^{-c}$$

and so the quantile function is:

$$Q(p|a,c) = a(1-p)^{-1/c}$$

---

## Example: Pareto Quantiles

So, to find the median of the Pareto distribution with $a=10$ and $c=4.5$, we can calculate:

```{r}
10*(1 - 0.5)^(-1/4.5)
```

For the 90th percentile:

```{r}
10*(1 - 0.9)^(-1/4.5)
```

If we want to change the scale parameter to 6 and the location parameter to 25:

```{r}
25*(1 - 0.9)^(-1/6)
```

However, re-submitting a slightly different calculation every time is tedious and prone to error. A better approach is to generalize the calculation in a re-usable function.

---

## Example: Pareto Quantiles

One possible implementation of the Pareto quantile function:

```{r}
qpareto <- function(p, scale, location) {
    q <- location * (1 - p)^(-1/scale)
    return(q)
}
```

Here, we have declared a function using `function` and assigned it to a variable, `qpareto`. We can now call the function with any combination of parameters, as required:

```{r}
qpareto(0.5, scale=4.5, location=10)
qpareto(0.9, scale=4.5, location=10)
qpareto(0.9, scale=6, location=25)
```

However, our function leaves room for improvement:

```{r}
qpareto(0.9, 6, -1)
```

---

## Error Checking

To make this function more robust, we can add some simple error checking statements that prevent illegal parameters from generating invalid results:

```{r}
  qpareto <- function(p, scale, loc) {
    if ((scale<=0) | (loc<=0)) {
        stop("'qpareto' parameters must be greater than zero.")
    }
    q <- loc*(1 - p)^(-1/scale)
    return(q)
}
```

Now, entering negative parameters raises an exception that must be dealt with, rather than an invalid result that could silently corrupt subsequent results.

```{r}
tryCatch(qpareto(0.4, 5, -1), error=function(e) e)
```

We could improve this further by checking to ensure that `p` is a valid probability.

---

## Error Checking

To avoid a long series of conditional statements, we can use a convenience function, `stopifnot`:

```{r}
qpareto <- function(p, scale, loc) {
    stopifnot(p>=0, p<=1, scale>0, loc>0)
    q <- loc*(1 - p)^(-1/scale)
    return(q)
}
```

`stopifnot` halts the execution of the function, with an error message, if all of its arguments do not evaluate to TRUE. If all those conditions are met, however, R just goes on to the next command.

```{r}
tryCatch(qpareto(-0.1, 4, 5), error=function(e) e)
tryCatch(qpareto(0.5, 0, 12), error=function(e) e)
```

## Presenter Notes

If the arguments violate two conditions, only the first one is cited.

---

## Function Scope

Variables defined inside a function, or in its arguments, are only available inside the function.

```{r}
(x <- 7)
square <- function(y) { x <- y^2; return(x) }
square(7)
x
tryCatch(y, error = function(e) e)
```

The assignment of 7 to `y` and of 49 to `x` only applies within the *scope* of the function. In other words, `y` is *local* to `square`, but `x` is *global*.

On the other hand, functions can see variables described in its environment:

```{r}
z <- 10
adder <- function(x) { return(x+z) }
adder(5)
```

## Presenter Notes

Relying on variables outside the function is bad practice

---

## Environments and Scoping

In the previous slide, both the variable `x` and the function `square` were created in the top-level environment, `.GlobalEnv`.

```{r, echo=FALSE}
# remove all variables except x and square (and metadata)
rm(list=setdiff(ls(), c('x','square','metadata')))
```

```{r}
ls()
ls.str()
environment(square)
```

Here's another example with a function `h` that is created inside of another function `f`. Its environment is referred to by its memory location:

```{r}
w <- 12
f <- function(y) {
   d <- 8
   h <- function() return(d*(w+y))
   print(environment(h))
   return(h())
}
f(2)
```

---

## Respect Interfaces

Interfaces allow us to completely control what is calculated inside a function.

Best practices:

* explicitly pass all external information required by a function via its arguments
* only allow a function to modify its environment through its return value

There are a few exceptions to these rules, such as universal constants and plotting behavior.

---

## Side Effects

When functions change non-local variables, this is known as a *side-effect*.

In general, side effects are considered poor practice because it is easy to produce unexpected results.

R makes it difficult to generate side effects because it makes local copies of all global variables when they are used inside a function.

```{r}
v <- 1:10
u <- function() { v[3] <- 0 }
u()
v
```

---

## Example: Fixed-point Iteration

Here is the algorithm for fixed-point iteration:

```{r}
# Initialize value
x <- 10
x_old <- x + 1e6
# Set tolerance
tol <- 1e-9
# Specify maximum number of iterations
max_iter <- 100
# Keep track of number of interations
iter <- 0
# Loop
while((abs(x-x_old) > tol) && (iter < max_iter)) {
    # Replace old value with current
    x_old <- x
    # Calculate new value
    x <- exp(exp(-x))
    # Increment counter
    iter <- iter + 1
}
x
```

---

## Encapsulate in a Function

To avoid cutting-and-pasting every time we want to re-run the code, we can make a function:

```{r}
fixedpoint <- function(x) {
    x_old <- x + 1e6
    # Set tolerance
    tol <- 1e-9
    # Specify maximum number of iterations
    max_iter <- 100
    # Keep track of number of interations
    iter <- 0
    # Loop
    while((abs(x-x_old) > tol) && (iter < max_iter)) {
        # Replace old value with current
        x_old <- x
        # Calculate new value
        x <- exp(exp(-x))
        # Increment counter
        iter <- iter + 1
    }
    return(x)
}
fixedpoint(10)
```

*That was easy!*

---

## Default Arguments

We may want to specify alternate tolerance or stopping time, so we can make these arguments with reasonable defaults:

```{r}
fixedpoint <- function(x, tol=1e-9, max_iter=100) {
    x_old <- x + 1e6
    # Keep track of number of interations
    iter <- 0
    # Loop
    while((abs(x-x_old) > tol) && (iter < max_iter)) {
        # Replace old value with current
        x_old <- x
        # Calculate new value
        x <- exp(exp(-x))
        # Increment counter
        iter <- iter + 1
    }
    return(x)
}
```

*Are we finished?*

---

## Generalization

We may wish to find the root of a different function, so lets allow it to be passed as an argument:

```{r}
fixedpoint <- function(fun, x=1, tol=1e-9, max_iter=100) {
    x_old <- x + 1e6
    # Keep track of number of interations
    iter <- 0
    # Loop
    while((abs(x-x_old) > tol) && (iter < max_iter)) {
        # Replace old value with current
        x_old <- x
        # Calculate new value
        x <- fun(x)
        # Increment counter
        iter <- iter + 1
    }
    return(x)
}

f <- function(x) exp(exp(-x))
f2 <- function(x) x - log(x) + exp(-x)
fixedpoint(f)
fixedpoint(f2)
```

*One more thing ...*

---

## Error-checking

We should ensure that the algorithm actually converged:

```{r}
fixedpoint <- function(fun, x=0, tol=1e-9, max_iter=100) {
    x_old <- x + 1e6
    # Keep track of number of interations
    iter <- 0
    # Loop
    while((abs(x-x_old) > tol) && (iter < max_iter)) {
        # Replace old value with current
        x_old <- x
        # Calculate new value
        x <- fun(x)
        # Increment counter
        iter <- iter + 1
    }

    # Check convergence
    if (abs(x-x_old) > tol) {
      cat("Algorithm failed to converge\n")
      return(NULL)
    } else {
      cat("Algorithm converged\n")
      return(x)
    }
}

fixedpoint(f)
fixedpoint(f, max_iter=3)
```

---

## Functions for Functions

There are a few additional techniques useful in the scope of writing functions.

* Global assignment: `assign()`, `"<<-"`

    Using these may violate the *respect interfaces* principle.

* The dots `...` argument

    Used to allow additional function arguments.  Typically this is used to pass arguments from one function to another - this is extremely useful for graphical parameters.  The components of `...` can also be stored with `list(...)`.

* `match.call()`

    Used to record a function call for later re-use.  Or you may be interested in printing out how a function was called with `deparse(match.call())`.

* `match.arg()`

    Matches an argument against candidate values.

* `missing()`

    Tests whether a value was specified as a function argument.

---

```{r}
myhist <- function(x, ...) {
  extArgs <- list(...)
  print(names(extArgs))
  title <- deparse(match.call())
  x <- hist(x, plot=FALSE)
  plot(NULL, xlim=range(x$breaks), ylim=range(x$counts), xlab='', ylab='', type='n', axes=FALSE, frame.plot=TRUE, main=title, ...)
  axis(1, at=x$breaks, labels=x$breaks)
  for(i in seq(length(x$counts))) {
    segments(x$mids[i], 0, x$mids[i], x$counts[i])
    text(x$mids[i], x$counts[i], x$counts[i], pos=4)
  }
}
myhist(islands, sub="Areas of major landmasses")

center <- function(x, type=c('mean','median')) {
  type <- match.arg(type)
  do.call(type, list(x))
}
set.seed(35)
x <- rnorm(10000)
center(x)
center(x, 'median')
center(x, 'med')
tryCatch(center(x, 'midpoint'), error=function(e) e)
```

---

## Exercise

### Implement the Newton-Raphson algorithm as a **function**
