---
title: "Data Types and Structures in R"
output: pdf_document
---

## Data Types

R has several elemental data types.

* numeric
* integer
* complex
* logical
* character

---

## Numeric

Floating-point numbers are represented by the `numeric` type in R. It is the default computational data type.

```{r}
x <- 10.5      # assign a decimal value
x
class(x)       # print the class name of x
```

Even if we assign an integer to a variable, it gets saved as a `numeric` type:

```{r}
k <- 1
k
class(k)       # print the class name of k
```

We can query this directly:

```{r}
is.integer(k)  # is k an integer?
```

---

## Integer

Integers can be created by *casting* a numeric value, using the `as.integer()` function or attaching the `L` suffix:

```{r}
y <- as.integer(3)
y
class(y)
is.integer(y)
is.integer(5L)
```

We can cast other types to integers this way:

```{r}
as.integer(42.1343)
as.integer("6.778")
as.integer(TRUE)
```

---

## Complex

A complex value in R is defined via the pure imaginary value i.

```{r}
z <- 1 + 2i
z
class(z)
```

The following gives a warning as -1 is not a complex value.

```{r}
sqrt(-1)
```

Instead, we have to use the complex value -1 + 0i.

```{r}
sqrt(-1+0i)
sqrt(as.complex(-1))
```

---

## Logical

A logical value is often created via comparison between variables.

```{r}
x <- 1; y <- 2
z <- x > y
z
class(z)
```

Standard logical operations are "&" (and), "|" (or), and "!" (negation).

```{r}
u <- TRUE; v <- FALSE
u & v          # u AND v
u | v          # u OR v
!u             # negation of u
```

---

## Character

A character object is used to represent string values in R. We convert objects into character values with the as.character() function:

```{r}
(x = as.character(3.14))
class(x)
```

Two character values can be concatenated with the paste function.

```{r}
fname = "Joe"; lname ="Smith"
paste(fname, lname)
```

---

## Vectors

The *fundamental* data structure in R.

A vector is a data structure composed of multiple elements, all of the same *mode*.

```{r}
x <- c(88,5,12,13)
x
mode(x)
```

The function `c` combines the elements specified into a vector.

Notice what happens when we try to change the data type of one of the elements:

```{r}
x[3] <- 'a'
mode(x)
```

---

### Exercise

You now know about variable types: character, logical, numeric, integer, complex.  Try creating different combinations of these types (by creating vectors).  For instance, combining a numeric and integer will create a numeric-type vector.  This is because numeric types are less flexible than integer types.  Determine the order of flexibility for all five.

```{r}
mode(c(1, 1L)) # integer is more flexible than numeric
```

Reflect on the following:

```{r}
1 == "1"
-1 < FALSE
"10" < "3"
"one" < 2
```

---

### Presenter Notes

The size of the vector is allocated to memory when it is created.

---

## Indexing

The elements of a vector can be accessed via *indexing*, using square brackets.

```{r}
x <- c(88,5,12,13)
x[2]
x[1:3]
x[c(4,1)]
x[-2]
x[x>10]
big <- which(x>10)
x[big]
```

Scalars are just one-element vectors:

```{r}
y <- 10
y[1]
y[2]
```

### Presenter Notes

which() takes a Boolean vector and gives a vector of indices for the TRUE values; useful with tests

---

## Vector Arithmetic

Operators apply to vectors "element-wise":

```{r}
y <- c(-7, -8, -10, -45)
x+y
```

**Recycling**: elements in shorter vector are repeated when combined with a longer vector

```{r}
x + c(-7,-8)
```

---

## Vector Comparisons

Can also do pairwise comparisons, which returns a boolean vector:

```{r}
x > 9
```

Boolean operators work pairwise; but written double, combines individual values into a single Boolean:

```{r}
(x > 9) & (x < 20)
(x > 9) && (x < 20)
```

To compare whole vectors, best to use `identical()` or `all.equal()`:

```{r}
x == y
identical(c(0.5-0.3,0.3-0.1),c(0.3-0.1,0.5-0.3))
all.equal(c(0.5-0.3,0.3-0.1),c(0.3-0.1,0.5-0.3))
```

---

## Special Values

In addition to the standard data types and data structures, there are four "special" values that you will encounter in R.

* `NA`
* `Inf`
* `NaN`
* `NULL`

---

## NA

`NA` is used to represent missing values (*i.e.* not available). For example, if you expand the length of a vector beyond the indices where values were defined, `NA` values are inserted:

```{r}
v <- c(1,2,3)
v
length(v) <- 4
v
```

---

## Inf

`Inf` represents infinity. It is most commonly encountered when a computation results in a number that is too big:

```{r}
2^1024
-2^1024
```

This is also the value returned when you divide by 0:

```{r}
1/ 0
```

---

## NaN

`NaN` stands for "not a number", and it is returned when a computation is nonsensical:

```{r}
Inf - Inf
0/0
```

---

## NULL

The `NULL` value indicates that there is no value assigned to a particular object. Note the difference between this and the notion of a missing value or a nonsensical value.

---

## Vectorized Functions

Lots of functions take vectors as arguments:

- `mean()`, `median()`, `sd()`, `var()`, `max()`, `min()`, `length()`, `sum()` yield single numbers
- `sort()` returns a new vector
- `hist()` takes a vector of numbers and produces a histogram and plots it
- `summary()` gives a five-number summary of numerical vectors
- `any()` and `all()` are useful on Boolean vectors

For example:

```{r}
mean(x)
sort(x)
summary(x)
any(x > 10)
```

### Presenter Notes

Making a plot is a "side-effect"

---

## Named Elements

You can give names to elements or components of vectors

```{r}
names(x) <- c("v1","v2","v3","foo")
names(x)
x[c("foo","v1")]
```

Labels are not actually part of the value;
`names(x)` is just another vector (of characters):

```{r}
names(y) <- names(x)
sort(names(x))
which(names(x)=="foo")
```

You can also remove the names, temporarily or permanently:

```{r}
unname(x)
x
names(x) <- NULL
x
```

---

## Arrays

Arrays add additional structure to vectors.

```{r}
x_arr <- array(x,dim=c(2,2))
x_arr
```

Notice the array is filled column-wise. Arrays have several properties:

```{r}
dim(x_arr)
is.vector(x_arr)
is.array(x_arr)
typeof(x_arr)
str(x_arr)
attributes(x_arr)
```

### Presenter Notes

- arrays can be n-dimensional.
- typeof() returns the type of the elements
- str() gives the structure: here, a numeric array

---

## Array Indexing

Can access a 2-D array either by pairs of indices or by the underlying vector:

```{r}
x_arr[1,2]
x_arr[3]
```

Omitting an index means "all of it":

```{r}
x_arr[c(1:2),2]
x_arr[,2]
```

---

## Operating on Arrays

Some functions do not respect array structure, and operate on the underlying vector instead:

```{r}
which(x_arr > 9)
```

Other functions preserve array structure:

```{r}
y_arr <- array(y,dim=c(2,2))
y_arr + x_arr
```

Others specifically act on each row or column of the array separately:

```{r}
rowSums(x_arr)
```

---

## Matrices

In R, a matrix is a specialization of a 2D array, with attributes for the numbers of columns and rows.

```{r}
dataset <- matrix(c(1,40,1,60),nrow=2)
dataset
is.array(dataset)
is.matrix(dataset)
(dataset2 <- matrix(c(1,40,1,60),nrow=2,byrow=T))
```

Matrices can be compared using functions like `identical` and `all.equal`:

```{r}
identical(dataset, dataset2)
all.equal(dataset, dataset2)
```

### Presenter Notes

Alternately, could specify `ncol`.
Element-wise operations with the usual arithmetic and comparison operators.

---

## Matrix Multiplication

Matrix multiplication has a special operator:

```{r}
(diag_3 <- diag(2)*3)
dataset2 %*% diag_3
```

Multiplying by a vector:

```{r}
y <- c(10,20)
dataset2 %*% y
y %*% dataset2
```

### Presenter Notes

R silently casts the vector as a row or column matrix.

---

## Matrix Operations

Matrix transpose:

```{r}
t(dataset2)
```

Matrix determinant:

```{r}
det(dataset2)
```

Extracting or replacing the diagonal:

```{r}
diag(dataset2)
diag(dataset2) <- c(4,12)
dataset2
```

## Matrix Operations

Creating a diagonal matrix or an identity matrix:

```{r}
diag(c(3,4))
diag(2)
```

Accessing the lower and upper triangular part of a matrix:

```{r}
m2 <- matrix(1:9, 3, 3)
upper.tri(m2)
m2[lower.tri(m2)] <- NA
m2
```

---

## Matrix Operations

Inverting a matrix:

```{r}
solve(dataset)
solve(dataset) %*% dataset # Just to check!
```

Solving the linear system $\mathbf{A}\mathbf{x} = \mathbf{b}$ for unknown $\mathbf{x}$:

```{r}
dataset
y
solve(dataset, y)
dataset %*% solve(dataset,y)
```

---

## Naming Rows and Columns

We can name either rows or columns or both, with rownames() and colnames():

```{r}
rownames(dataset2) <- c("subject1", "subject2")
colnames(dataset2) <- c("intercept", "dose")
dataset2
```

### Presenter Notes

These are just character vectors, and we use the same function to get and to set their values
Names help us understand what we’re working with

---

## Lists

Lists allow for the storage of a sequence of values, not necessarily of the same type:

```{r}
my_distribution <- list("exponential",7,FALSE)
my_distribution
```

Most of things which you can do with vectors you can also do with lists!

---

## Indexing Lists

Use single square brackets (*i.e.* `[]`) to index a sub-list and double square brackets (*i.e.* `[[]]`) to access the list element.

```{r}
class(my_distribution)
class(my_distribution[1])
class(my_distribution[[1]])
```

What happens when you index a vector with double brackets?

---

## List Manipulation

Add to lists with `c()` (also works with vectors):

```{r}
my_distribution <- c(my_distribution,7)
my_distribution
```

Chop off the end of a list by setting length to something smaller:

```{r}
length(my_distribution)
length(my_distribution) <- 3
my_distribution
```

---

## Named List Elements

Lists are most effective when the list items are named:

```{r}
another_distribution <- list(family="gaussian",mean=7,sd=1,is.symmetric=TRUE)
another_distribution
```

This allows access by name:

```{r}
another_distribution$family
another_distribution["family"]
```

### Presenter Notes

Lists give us a way to store and look data up by name rather than number (key-value pairs, dictionary, associative array, hash)

---

## Named List Elements

Similarly, list elements can be added by name:

```{r}
another_distribution$was_estimated <- FALSE
another_distribution
```

Remove an entry in the list by assigning it the value NULL.

```{r}
another_distribution$was.estimated <- NULL
```

---

## Data Frames

The `data.frame` is an analog of a spreadsheet data table, with `n` rows, representing observations or cases, and `p` columns, representing variables.

`data.frame` generalizes both the `matrix` and `list`

- columns are named
- columns can be of different types
- many matrix functions work: `rowSums()`, `summary()`
- no matrix operations (*e.g.* multiply, transpose)

The "default" data strucuture for statistical operations in R.

---

## Data Frames

First, consider a `matrix` object, which is another sort of tabular structure we have seen.

```{r}
(data_matrix <- matrix(c(55.5,69.5,1,41,81.5,1,53.5,86,1), ncol=3, byrow=T))
colnames(data_matrix) <- c("test1","test2","disease")
data_matrix
```

---

## Data Frames

We can *cast* the matrix to a `data.frame` as follows:

```{r}
df <- data.frame(data_matrix)
df
```

and easily manipulate the columns/rows:

```{r}
df$disease <- as.logical(df$disease)
df
df['test2']
df[2,]
```

---

## Manipulating Data Frames

We can add rows or columns to an array or data-frame with `rbind()` and `cbind()`.

```{r}
rbind(df, list(test1=55.1, test2=80.4, disease=FALSE))
rbind(df, c(55.1, 80.4, 2))
cbind(df, list(sex=c('M','F','F')))
```

### Presenter Notes

be careful about forced type conversions

---

## Structures of Structures

Arbitrarily-complex data structures can be constructed from basic ones.

- lists of lists
- lists of vectors
- lists of lists of lists of vectors

Recursion allows us to build them.

---

## Example: eigenstructure

`eigen()` calculates eigenvalues and eigenvectors of a matrix, and places them in a list:

```{r}
eigen(dataset2)
class(eigen(dataset2))
str(eigen(dataset2))
dataset2 %*% eigen(dataset2)$vectors[,2]
eigen(dataset2)[[1]][[2]]
```

### Exercise

Create a list with the following named items, and assign it to `x`:

* data = 1:20
* nrow = 5
* byrow = TRUE

```{r}
x <- list()
```

Run the following, while making the requested modifications.

```{r}
do.call(matrix, x)
```

Modify the data element to contain `sample(20)`.  Remove the `byrow` element.

```{r}
do.call(matrix, x)
```

Remove the `nrow` element and add an element called `ncol` with a value of 2.

```{r}
do.call(matrix, x)
```
