---
title: "R data structures"
output: html_notebook
---

# R data types
* logical
* integer
* numeric
* complex
* character

Reflect on the following:

```{r}
1 == "1"
-1 < FALSE
"10" < "3"
"one" < 2
```

# Special values
* `NA`
* `Inf`
* `NaN`
* `NULL`

```{r}
2 ^ 1024
(-2) ^ 1024
-2 ^ 1024
1 / 0
Inf - Inf
0 / 0
is.infinite(1 / 0)
```

# R data structures (containers)
* vector
* matrix
* array
* data.frame
* list

# Creating a Vector

```{r}
numeric(10)
rep(1, 10)
rep(1:4, times = 1:4)
rep(seq(4), each=4)
(x <- seq(10, 100, by = 10))
x[15] <- 150
x
```

# Vector Functions

Some include: head, rev, match, %in%, sort, order, table, unique, duplicated, union, intersect, setdiff, `[`

```{r}
x <- c(88, 5, 12, 13)
x[2]
x[1:3]
x[c(4,1)]
x[-2]
x[x > 10]
big <- which(x > 10)
x[big]
table(sample(x, 100, replace=TRUE))
```

```{r}
y <- c(12, 20, 88)
# find first element of y in x
which(y[1] == x)
# find all of y in x: BAD
which(y == x)
# find all of y in x: GOOD
match(y, x)
# match finds first occurrence only
match(y, c(x, x))
y %in% x
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
duplicated(c(x, y))
```

The `names` attribute

```{r}
names(x)
names(x) <- c("v1", "v2", "v3", "foo")
x
x[c("foo", "v1")]
sort(names(x))
which(names(x) == "foo")
unname(x)
x
names(x) <- NULL
x
```

# Creating a Matrix

```{r}
x <- seq(16)
matrix(x)
matrix(x, nrow = 4)
# matrix(x, 4)
matrix(x, ncol = 8, byrow = TRUE)
diag(4)
```

# Matrix Functions

Some include: colSums, colMeans, cbind, t, diag, upper.tri, lower.tri, solve, cor, `[`, `%*%`

```{r}
x <- seq(16)
y <- rep(1:4, each=4)
x * y
x %*% y
matrix(x) %*% y
x %*% matrix(y)
matrix(x) %*% matrix(y)
```

```{r}
mx <- matrix(x, 4)
my <- matrix(y, 4)
mx * my
mx %*% my
my %*% mx
```

