---
title: "R data structures"
output: html_notebook
---

# R data types
* logical
* integer
* numeric
* complex
* character

Reflect on the following:

```{r}
1 == "1"
-1 < FALSE
"10" < "3"
"one" < 2
```

# Special values
* `NA`
* `Inf`
* `NaN`
* `NULL`

```{r}
2 ^ 1024
(-2) ^ 1024
-2 ^ 1024
1 / 0
Inf - Inf
0 / 0
is.infinite(1 / 0)
```

# R data structures (containers)
* vector
* matrix
* array
* data.frame
* list

# Creating a Vector

```{r}
numeric(10)
rep(1, 10)
rep(1:4, times = 1:4)
rep(seq(4), each=4)
(x <- seq(10, 100, by = 10))
x[15] <- 150
x
```

# Vector Functions

Some include: head, rev, match, %in%, sort, order, table, unique, duplicated, union, intersect, setdiff, `[`

```{r}
x <- c(88, 5, 12, 13)
x[2]
x[1:3]
x[c(4,1)]
x[-2]
x[x > 10]
big <- which(x > 10)
x[big]
table(sample(x, 100, replace=TRUE))
```

```{r}
y <- c(12, 20, 88)
# find first element of y in x
which(y[1] == x)
# find all of y in x: BAD
which(y == x)
# find all of y in x: GOOD
match(y, x)
# match finds first occurrence only
match(y, c(x, x))
y %in% x
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
duplicated(c(x, y))
```

The `names` attribute

```{r}
names(x)
names(x) <- c("v1", "v2", "v3", "foo")
x
x[c("foo", "v1")]
sort(names(x))
which(names(x) == "foo")
unname(x)
x
names(x) <- NULL
x
```

# Creating a Matrix

```{r}
x <- seq(16)
matrix(x)
matrix(x, nrow = 4)
# matrix(x, 4)
matrix(x, ncol = 8, byrow = TRUE)
diag(4)
```

# Matrix Functions

Some include: colSums, colMeans, cbind, t, diag, upper.tri, lower.tri, solve, cor, `[`, `%*%`

```{r}
x <- seq(16)
y <- rep(1:4, each=4)
x * y
x %*% y
matrix(x) %*% y
x %*% matrix(y)
matrix(x) %*% matrix(y)
```

```{r}
dataset <- matrix(c(1,40,1,60), nrow=2)
rownames(dataset) <- c('intercept', 'dose')
colnames(dataset) <- c('subject1', 'subject2')
dataset
t(dataset)
solve(dataset)
solve(dataset) %*% dataset
y <- c(10, 20)
solve(dataset, y)
dataset %*% solve(dataset, y)
```

`mx` and `my` and 2x2

```{r}
mx <- matrix(x, nrow=4)
my <- matrix(y, nrow=4)
mx * my
mx %*% my
my %*% mx
```

`mx1` is 8x2 and `my1` and 2x8

```{r}
mx1 <- matrix(x, ncol=2)
my1 <- matrix(y, nrow=2)
mx1 %*% my1
my1 %*% mx1
```

`mx2` is 2x8 and `my2` and 8x2

```{r}
mx2 <- matrix(x, ncol=8, byrow=TRUE)
my2 <- matrix(y, nrow=8)
mx2 %*% my2
my2 %*% mx2
t(mx2) %*% t(my2)
mx2 %*% t(my1)
```

```{r}
mx2 %*% diag(8)
mx * upper.tri(mx)
mx * lower.tri(mx, diag=TRUE)
```

Explain the following

```{r}
(x <- matrix(seq(25), 5, 5))
x[x %% 2 == 0] <- NA
x[upper.tri(x, diag=TRUE)] <- NA
x[,1] <- x[5,] <- NA
x
which(!is.na(x), arr.ind=TRUE)
```

The complicated one-liner

```{r}
which(!(x %% 2 == 0 | upper.tri(x, diag=TRUE) | (x-1) %/% 5 == 0 | x %% 5 == 0), arr.ind=TRUE)
```
