---
title: "R data structures"
output: html_notebook
---

# Creating a data.frame

```{r}
y <- data.frame(id=1:3, animal=c('cat','dog','rabbit'), 
                speed=c('faster','fast','fastest'), stringsAsFactors=FALSE)
y[,'speed.factor'] <- factor(y[,'speed'])
y
```

# Extracting a column

```{r}
y[,3]
y[,'speed']
y$speed
y[['speed']]
y[,'speed', drop=FALSE]
y['speed']
```

# Adding/removing a column

```{r}
y[,'new'] <- NA
y <- cbind(y, newer=0)
y
y[,'new'] <- NULL
y <- y[,-1]
y <- y[,-match('newer', names(y))]
y
```

# Data.frame functions

Some include: merge, with, within, & many of the matrix functions (cbind, colSums)

```{r}
x <- data.frame(id = seq(4, 24, 4), gender = rbinom(6, 1, 0.5))
y <- data.frame(id = seq(6, 24, 6), gender = rbinom(4, 1, 0.25))
merge(x, y)
merge(x, y, by=0)
merge(x, y, by.x='id', by.y='id', all.x=TRUE, all.y=FALSE)
merge(x, y, all=TRUE)
```

```{r}
x <- within(x, {
  weight <- round(rnorm(nrow(x), 120 + gender * 60, 10))
  age <- sample(15:25, nrow(x), replace = TRUE)
})
x[order(x[,'gender'], x[,'age']),]
x[with(x, order(gender, age)),]
```

### Exercise

1. Create a 5x5 matrix with the first 25 odd numbers.
2. If a number is divisible by 3, replace it with NA.
3. Use the `colMeans` function to determine the mean of each column (hint: use `na.rm` argument).
4. Replace missing values with mean of the value from the column in which they are found.
5. Use `as.data.frame` to convert your matrix into a data.frame

```{r}
```

# Creating a list

```{r}
my_dist <- list("exponential", 7, FALSE)
my_dist
my_dist[1]
my_dist[[1]]
```

```{r}
my_dist <- c(my_dist, 7)
my_dist[[6]] <- list("apple", "banana")
my_dist
```

```{r}
another_dist <- list(family = "gaussian", mean = 7, sd = 1, is.symmetric = TRUE)
another_dist
another_dist$family
another_dist[['family']]
another_dist$was_estimated <- FALSE
another_dist$was_estimated <- NULL
```

### Exercise

Create a list with the following named items, and assign it to `x`:

* data = 1:20
* nrow = 5
* byrow = TRUE

```{r}
x <- list()
```

Run the following, while making the requested modifications.

```{r}
do.call(matrix, x)
```

Modify the `data` element to contain `sample(20)`.  Remove the `byrow` element.

```{r}
do.call(matrix, x)
```

Remove the `nrow` element and add an element called `ncol` with a value of 2.

```{r}
do.call(matrix, x)
```
