---
title: "Control Structures"
output: html_notebook
---

# Conditional code execution

## `if` statements

```{r}
x <- -10
if (x >= 0) {
    x
} else {
    -x
}
```

How is truth evaluated?

```{r}
if(1) {"Truth!"} else {"Falsehood!"}
if(-1) {"Truth!"} else {"Falsehood!"}
if(0) {"Truth!"} else {"Falsehood!"}
if("TRUE") {"Truth!"} else {"Falsehood!"}
tryCatch(if("TRUTH") {"Truth!"} else {"Falsehood!"}, error=function(e) e)
tryCatch(if("c") {"Truth!"} else {"Falsehood!"}, error=function(e) e)
tryCatch(if(NULL) {"Truth!"} else {"Falsehood!"}, error=function(e) e)
tryCatch(if(NA) {"Truth!"} else {"Falsehood!"}, error=function(e) e)
tryCatch(if(NaN) {"Truth!"} else {"Falsehood!"}, error=function(e) e)
```

## Adding `else if` and `else`

```{r}
age <- 10
if (age < 2) {
    group <- 'infant'
} else if (age < 12) {
    group <- 'child'
} else if (age < 18) {
    group <- 'adolescent'
} else {
    group <- 'adult'
}
group
```

## Nested conditionals

```{r}
if (x^2 < 1) {
    x^2
} else {
    if (x >= 0) {
        x
    } else {
        -x
    }
}
```

In the above, `x` is 10.  What would happen if x where a vector?

## `switch` statement

```{r}
centre <- function(x, type) {
   switch(type,
          mean = mean(x),
          median = median(x),
          trimmed = mean(x, trim = .1))
 }
x <- rnorm(10)
centre(x, "mean")
centre(x, "median")
centre(x, "trimmed")
```

# Code Iteration

## `for` loop

```{r}
x <- c(5,12,13)
for(n in x) print(n^2)
```

```{r}
n <- 10
logs <- numeric(n)
for(i in 1:n) {
  logs[i] <- log(i)
}
logs
```

```{r}
x <- c(-5,7,-8,0)
y <- rep(NULL, length(x))
for (i in 1:length(x)) {
    if (x[i] >= 0) {
        y[i] <- x[i]
    } else {
        y[i] <- -x[i]
    }
}
y
```

## `while` loop

```{r}
i <- 0
while(i < 2) i <- i + rnorm(1)
i
```

```{r}
i <- 0
while(TRUE) {
  i <- i+1
  if (i > 10) break
}
i
```

```{r}
x <- 0
repeat {
    x <- x + log(i)
    if (x > 20) break
}
x
```

```{r}
a <- rnorm(15)
b <- rnorm(15)
c <- numeric(length(a))
for (i in seq_along(a)) {
    c[i] <- a[i] + b[i]
}
c
# vectorized
a + b
# not vectorized
pmin(a,b)
```

```{r}
x <- rnorm(5, 0, 25)
abs(x)
log(x)
ifelse(x < 0, -x, x)
ifelse(x^2 > 1, abs(x), x^2)
```

# Exercise

Generate a vector of the first 25 prime numbers

```{r}

```

## `expand.grid`

```{r}
expand.grid(intervention=c("control", "treatment"), gender=c("M","F"))
```

## `outer`

```{r}
outer(c(1,3,5),c(2,3,7),'*')
```

## `replicate`

```{r}
mean(replicate(1000, sd(rnorm(1000))))
```

## Using `apply`s

```{r}
x <- expand.grid(intervention=c("control", "treatment"), gender=c("M","F"))
x[,'los'] <- rpois(4, 5)
x[,'smoker'] <- rbinom(4, 1, 0.5)
x
split(x, x[,'intervention'])
```

```{r}
apply(x[,c('los','smoker')], MARGIN = 2, FUN = sd)
```

In the above, why is MARGIN = 2?

```{r}
lapply(split(x[,'los'], x[,'intervention']), mean)
```

```{r}
sapply(split(x[,'los'], x[,'intervention']), mean)
```

```{r}
tapply(x[,'los'], x[,'intervention'], mean)
```

```{r}
vapply(split(x[,'los'], x[,'intervention']), mean, numeric(1))
```

```{r}
mapply(min, a, b)
# pmin(a, b)
```

# Exercise

Implement the Newton-Raphson algorithm in R

## Find root for $f(x) = x^3 + x^2 - 3x - 3$. $f'(x) = 3x^2 + 2x - 3$.

```{r}

```
