---
title: "Functions"
output: html_notebook
---

Define a function (DRY)

```{r}
my_function <- function(first_argument, second_argument, ...) {
  # do some stuff
  return(return_value)
}
```

Using `return` is optional.

```{r}
absval <- function(value) {
    if (value < 0) {
        -value
    } else {
        value
    }
}
absval(-5)
```

## Operators are functions too

```{r}
'+'(4,5)
```

Users can write custom operators.

```{r}
"%!%" = function(x,y) x/sqrt(y)
6 %!% 7
```

## Exercise

Remember the `%in%` function?  Write a `%nin%` function.

## Default Argument Values

```{r}
psi <- function(x, c=1) {
  if(x^2 > c^2) {
    c * abs(x)
  } else {
    x^2
  }
}
z <- 0.5
psi(z)
```

```{r}
identical(psi(z, c=1), psi(z))
identical(psi(x=z, c=2), psi(c=2, x=z))
```

## Validating Input

Function to compute percentile of the Pareto distribution with location and scale parameters

```{r}
qpareto <- function(p, scale, location) {
  location * (1 - p) ^ (-1 / scale)
}
qpareto(0.5, 4.5, 10)
qpareto(0.9, 4.5, 10)
qpareto(0.9, 6, -1)
```

```{r}
qpareto <- function(p, scale, loc) {
  if((scale<=0) || (loc<=0)) {
      stop("'qpareto' parameters must be greater than zero.")
  }
  location * (1 - p) ^ (-1 / scale)
}
tryCatch(qpareto(0.4, 5, -1), error=function(e) e)
```

```{r}
qpareto <- function(p, scale, loc) {
  stopifnot(p >= 0, p <= 1, scale > 0, loc > 0)
  location * (1 - p) ^ (-1 / scale)
}
tryCatch(qpareto(-0.1, 4, 5), error=function(e) e)
tryCatch(qpareto(0.5, 0, 12), error=function(e) e)
```

## Functions creating functions

```{r}
nd <- function(m = 0, sd = 1) function(n) rnorm(n, m, sd)
nd()(5)
distr <- nd(m = 100, sd = 10)
distr(5); distr(10)
```

## Scoping

```{r}
w <- 12
f <- function(y) {
   d <- 8
   h <- function() d*(w+y)
   return(h())
}
f(2)
w <- 13
f(2)
d <- 4
f(2)
```

```{r}
ctr <- 5
counter <- function() {
  ctr <- 0
  f <- function() {
    ctr <<- ctr + 1
    cat("value of counter is", ctr, "\n")
  }
}
c1 <- counter()
c2 <- c3 <- counter()
c1()
c2()
c2()
c2()
c3()
ctr
```

## Example: Fixed-point Iteration

Solve the root of `f(x) = log(x) - exp(-x)`

```{r}
f <- function(x) log(x) - exp(-x)
plot(f, xlim = c(0, 15))
```

We can do this with fixed-point iteration:

```{r}
# Initialize value
x <- 10
x_old <- x + 1e6
# Set tolerance
tol <- 1e-9
# Specify maximum number of iterations
max_iter <- 100
# Keep track of number of interations
iter <- 0
# Loop
while((abs(x-x_old) > tol) && (iter < max_iter)) {
    # Replace old value with current
    x_old <- x
    # Calculate new value
    x <- exp(exp(-x))
    # Increment counter
    iter <- iter + 1
}
x
```

## Exercise

Write a robust function for fixed point iteration.

## Functions for functions

There are a few additional techniques useful in the scope of writing functions.

* Global assignment: `assign()`, `"<<-"`
* The dots `...` argument
* `match.call()`
* `match.arg()`
* `missing()`

```{r}
resample <- function(x, ...) {
  xx <- sample(x, replace = TRUE)
  print(quantile(xx, ...))
  xx
}
resample(seq(100))
resample(seq(100), probs = 0.5)
```

```{r}
center <- function(x, type=c('mean','median')) {
  type <- match.arg(type)
  do.call(type, list(x))
}
set.seed(35)
x <- rnorm(10000)
center(x)
center(x, 'median')
center(x, 'med')
tryCatch(center(x, 'midpoint'), error=function(e) e)
```
