---
title: "Data Import and Export in R"
output: html_notebook
---

```{r}
cat("Happy Tuesday!", "\n")
iterations <- 10
for (j in 1:iterations) {
  cat(j, "...")
  if (j==iterations) cat("Done!", "\n")
}
```

```{r}
some_file <- tempfile()
cat(file=some_file, "foo", "bar", seq(1:10), sep="\n")
some_file
```

## Exporting a matrix

```{r}
my_mat <- outer(c(1,4,6),c(8,-1,4))
# write(my_mat, "my_mat.dat", ncolumn=3)
write(my_mat, stdout(), ncolumn=3)
MASS::write.matrix(my_mat, stdout())
```

```{r}
write.table(my_mat, stdout())
```

## Exporting Data Frames

Read documentation on `write.table` - many arguments

```{r}
x <- data.frame(id=seq(10), height=round(rnorm(10,70,4)), weight=round(rnorm(10,180,20)*2)/2)
write.table(x, file = "foo.csv", sep = ",", col.names = NA)
write.table(x, file = stdout(), sep = ",", col.names = NA)
```

Note that `write.table` is inefficient for very large matrices; `MASS::write.matrix` is more memory-efficient.

## Importing two-dimensional data

See the function arguments for `read.table`

* Header line `header = FALSE`
* Column separator `sep = ""`
* Quoting strings `quote = "\"'"`
* Missing values `na.strings = "NA"`
* Unfilled lines `fill = !blank.lines.skip`
* White space in character fields `strip.white = FALSE`
* Blank lines `blank.lines.skip = TRUE`
* Classes for the variables `colClasses = NA`
* Comments `comment.char = "#"`
* Rows to read `skip = 0` and `nrows = -1`
* Character VS factor `stringsAsFactors = default.stringsAsFactors()`

```{r}
read.table("foo.csv", header = TRUE, sep = ",", row.names=1)
```

## `read.table` helper functions

    !r
    read.csv(file, header = TRUE, sep = ',', quote='\"', dec='.', fill = TRUE, ...)

    read.delim(file, header = TRUE, sep = '\t', quote='\"', dec='.', fill = TRUE, ...)

## Fixed-width format

```{r}
fixed_data <- tempfile()
cat(file=fixed_data, "1 -14  23", "1   7 -10", sep="\n")
read.fwf(fixed_data, width=c(1,4,4))
```

---

## scan

Everything uses `scan`, which can be called directly.

```{r}
cat("2 3 5 7", "11 13 17 19", file="ex.dat", sep="\n")
scan(file="ex.dat")
scan(file="ex.dat", what=list(x=0, y="", z=0), flush=TRUE)
```

`scan` can also be used for rudimentary data entry - or utilize `stdin()`

```{r, eval = FALSE}
x <- scan()
y <- read.csv(stdin(), header = FALSE)
```

`file` can also be a URL

```{r}
# play intramural softball
tryCatch(read.csv('http://data.vanderbilt.edu/~graywh/intramurals/softball/batting_stats_career.txt',
                  row.names=1, strip.white=TRUE)["Cole",], error=function(e) e)
```

## Importing Source Code

```{r, eval = FALSE}
source("my_script.R")
source("useful_functions.R")
```

```{r}
source("https://github.com/couthcommander/misc/raw/master/R/print.R")
prettyprint(my_mat)
```

## Redirecting Output

* `sink`
* `capture.output`

```{r}
sink("sink-examp.txt")
i <- 1:10
outer(i, i, "*")
sink() # stops redirection
system("cat sink-examp.txt")
```

```{r}
counts <- c(18,17,15,20,10,20,25,13,12)
outcome <- gl(3,1,9)
treatment <- gl(3,3)
capture.output(glm(counts ~ outcome + treatment, family = poisson()), file=NULL)
```

## Storing Data

Compressed and efficient

```{r}
x <- 1:4
save(x, file="x.RData") # encode
rm(x)
load("x.RData") # decode
x
```

## Importing from Other Systems

The package `foreign` provides import facilities for files produced by:

* **Minitab** read.mtp
* **S-PLUS** read.S
* **SAS** read.xport,read.ssd
* **SPSS** read.spss
* **Stata** read.dta

Many other R packages can be useful for different types of files

* database packages: RODBC, DBI, sqlite
* SPSS/Stata/SAS: haven
* Excel: readxl
* Python: feather
* GIS: rgdal
* large CSVs: `data.table::fread`
* memory-mapped: mmap, ff
* HTML: XML

Or, check out [rio](https://github.com/leeper/rio)

## Functions for Files and Paths

```{r}
getwd()
file <- file.path("~", "Documents", "junk091814.txt")
file
file.create(file)
dirname(file)
basename(file)
file.remove(file)
setwd("~")
head(list.files())
```

```{r echo=FALSE, results='hide'}
# clean-up files created during session
file.remove('foo.csv', 'ex.dat', 'sink-examp.txt', 'x.RData')
```

## Example

```{r}
cat(file="file1.dat", 5, 12, 13, sep=',')
cat(file="file2.dat", 7, 6, 1, sep=',')
cat(file="file3.dat", 14, 5, 5, sep=',')
list.files(pattern = "\\.dat$")
```

Can we write code to sum the contents of all three files?

```{r}
sumfiles <- function(dname = '.') {
  tot <- 0
  for(f in list.files(dname, pattern = "\\.dat$")) {
    tot <- tot + sum(scan(f, sep=',', quiet=TRUE))
  }
  tot
}
sumfiles()
```

```{r echo=FALSE, results='hide'}
# clean-up files created during session
file.remove('file1.dat', 'file2.dat', 'file3.dat')
```
