---
title: "Working with Dates and Times"
output: html_notebook
---

## `Date` Class

Internally, `Date` stores time as an integer value, which represents the number of days since January 1, 1970

```{r}
as.Date("2012-09-30")
class(Sys.Date())
unclass(Sys.Date())
as.Date("09/30/2012", format="%m/%d/%Y") # Custom formatting
as.Date(41180, origin="1900-01-01") # Choose a different origin
```

```{r}
Sys.Date() - as.Date("2003-05-17")
difftime(Sys.Date(), as.Date("2003-05-17"), units="secs")
```

## Date-Time Data

* `POSIXct`, calendar time, stores time as the number of *seconds* since the origin
* `POSIXlt`, local time,  stores a list of time attributes, which can be indexed

```{r}
unclass(Sys.time())
unclass(as.POSIXlt(Sys.time()))
```

## Date-Time Formatting

See `strptime`

```{r}
as.POSIXct("080406 10:11", format = "%y%m%d %H:%M")
as.POSIXct("2008-04-06 10:11:01 PM", format = "%Y-%m-%d %I:%M:%S %p")
as.POSIXct("08/04/06 22:11:00", format = "%m/%d/%y %H:%M:%S")
```

```{r}
dt <- as.POSIXct("080406 10:11", format = "%y%m%d %H:%M")
format(dt, "%m/%d/%Y %I:%M %p")
as.character(dt, format = "%m-%d-%y %H:%M")
```

## Exercise

Write a function to calculate the time at which you turn a given number of seconds old, defaulting to a billion:

```{r}
# bday_in_secs("1987-10-19 14:00:00")
```

## Complications

Sometimes R will change date-time classes on you without warning!

```{r}
dts <- data.frame(day = c("20081101", "20081101", "20081101", "20081101", "20081101",
        "20081102", "20081102", "20081102", "20081102", "20081103"),
    time = c("01:20:00", "06:00:00", "12:20:00", "17:30:00", "21:45:00", "01:15:00",
        "06:30:00", "12:50:00", "20:00:00", "01:05:00"),
    value = c(5, 5, 6, 6, 5, 5, 6, 7, 5, 5))
dts1 <- paste(dts$day, dts$time)
dts2 <- as.POSIXct(dts1, format = "%Y%m%d %H:%M:%S")
dts3 <- as.POSIXlt(dts1, format = "%Y%m%d %H:%M:%S")
dts_all <- data.frame(dts, ct = dts2, lt = dts3)
str(dts_all)
```

```{r}
dts_all <- dts
dts_all[['ct']] <- dts2
dts_all[['lt']] <- dts3
str(dts_all)
```

Rounding date-times can also result in casting to a different type:

```{r}
# bad
dts_all[, "ct"] <- round(dts_all[, "ct"], units = "hours")
class(dts_all[, "ct"])
```

```{r}
# good
dts_all[, "ct"] <- as.POSIXct(round(dts2, units = "hours"))
```

```{r}
# bad
dts_all[, "lt"] <- round(dts3, units = "hours")
dts_all[, "lt"]
```

```{r}
# good
dts_all[['lt']] <- round(dts3, units = "hours")
str(dts_all)
```

## Time Zones and DST

```{r}
(time1 <- dts_all$lt[5])
(time2 <- dts_all$lt[7])
while (time1 < time2) {
  # Increment 1 hour until they are equal
  time1$hour <- time1$hour + 1
  print(unlist(time1))
}
```

```{r}
print(sprintf("%s -- %s", time1, time2))
time1 == time2
time1
time2
```

Converting the date classes clears up the problem, as does concatenating the dates:

```{r}
as.POSIXlt(as.POSIXct(time1))
c(time1, time2)
```

## Specifying Time Zones

Universal time (UTC) is nice

```{r}
round(as.POSIXlt(dts1, format = "%Y%m%d %H:%M:%S", tz = "America/Chicago"), units = "hours")
round(as.POSIXlt(dts1, format = "%Y%m%d %H:%M:%S", tz = "CST"), units = "hours")
(dts4 <- round(as.POSIXlt(dts1, format = "%Y%m%d %H:%M:%S", tz = "UTC"), units = "hours"))
```

## Example

```{r}
mydata_lt <- data.frame(date = NA, value = dts_all$value)
mydata_lt$date <- dts_all$lt
mydata_lt
```

Let's say we actually want an entry every hour, with the value filled with the last observed value. *How would we do this?*

## Lubridate

```{r}
hasPackage <- tryCatch(find.package('lubridate'), error=function(e) e)
if(inherits(hasPackage, "error")) install.packages('lubridate')
library(lubridate)
```

```{r}
ymd("20110604")
mdy("06-04-2011")
dmy("04/06/2011")
```

Data with time information as well as dates can similarly be accomodated:

```{r}
ymd_hms("2012-09-29 12:00:00", tz="America/Chicago")
```

## Manipulating Dates and Times

```{r}
(ldate <- mdy_hms("12/31/2012 23:59:59"))
ldate + dseconds(1)
month(ldate) <- 8
ldate
```

```{r}
second(ldate)
tz(ldate)
yday(ldate)
wday(ldate)
wday(ldate, label=TRUE)
```

## Periods vs Durations

* period: relative to timeline
* duration: precise

```{r}
minutes(5)
dminutes(5)
```

```{r}
leap_year(2011)
ymd(20110101) + dyears(1)
ymd(20110101) + years(1)

leap_year(2012)
ymd(20120101) + dyears(1)
ymd(20120101) + years(1)
```

## Sequences and intervals

```{r}
(meetings <- now() + weeks(0:5))
```

```{r}
holiday <- interval(ymd("2017/11/19"), ymd("2017/11/25"))
meetings %within% holiday
```
