---
title: "String Manipulation"
output: html_notebook
---

We've previously discussed

* nchar
* toupper
* substr
* strsplit
* chartr
* charmatch

## `paste`

```{r}
i <- 4; paste("the square of", i, "is", i^2)
i <- 8; paste("the square of ", i, " is ", i^2, sep = "")
i <- 16; paste0("the square of ", i, " is ", i^2)
i <- 32; paste(c("the square of", i, "is", i^2), collapse = " ")
```

The optional argument sep can be used to put something other than a space between the pieces being spliced together.

## Exercise

* Create a data frame with `ID|SITE|VALUE`
  * ID - randomly generated numeric value
  * SITE - randomly generated character string with length 4
  * VALUE - random numeric value
* Create a column `key` by pasting together ID and SITE
* What if you had many columns and needed to paste them all together?

## `sprintf`

```{r}
sprintf("the square of %d is %d", i, i^2)
```

There are a variety of format options:

```{r}
sprintf("%.3f", pi)
sprintf("%1.0f", pi)
sprintf("%10f", pi)
sprintf("%010f", pi)
sprintf("%-10f", pi)
sprintf("%E", pi)
sprintf("%+f", pi)
```

## Exercise

Use `strsplit`, etc to reorder the following format `LAST, FIRST, POSITION` to `F. LAST (POSITION)`:

```{r}
lineup <- c(
"Springer, George, CF",
"Bregman, Alex, 3B",
"Altuve, Jose, 2B",
"Correa, Carlos, SS",
"Gurriel, Yuli, 1B",
"McCann, Brian, C",
"Gonzalez, Marwin, LF",
"Reddick, Josh, RF",
"Verlander, Justin, P"
)
```

## Finding string matches

```{r}
grep("test", c("foo", "bar", "test"))
grep("test", c("foo", "bar", "test"), value=TRUE)
grepl("test", c("foo", "bar", "test"))
```

## POSIX Character Classes

* [:alnum:] : any alphanumeric character 0 to 9 or A to Z or a to z.
* [:word:] : alphanumeric characters plus `_`
* [:alpha:] : any alpha character A to Z or a to z.
* [:digit:] : only the digits 0 to 9
* [:blank:] : space and tab characters only.
* [:graph:] : exclude whitespace (SPACE, TAB).
* [:lower:] : any alpha character a to z.
* [:print:] : any visible characters and the space character
* [:punct:] : any punctuation characters.
* [:upper:] : any alpha character A to Z.

```{r}
grep("[[:alpha:]]", c("----x----","--------","5456655"))
grep("[[:blank:]]", c("foo","bar","foo bar","foobar"))
```

## Substituting Text

```{r}
text <- c("arm", "leg", "head", "foot", "hand", "hindleg", "elbow")
sub("o", "O", text)
gsub("o", "O", text)
```
